---
title: "R-workshop"
author: 
- Foon Wong
- fwong@csu.fullerton.edu
date: "April 14, 2018"
output:
  ioslides_presentation: 
    widescreen: TRUE
    incremental: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE)
library(tidyverse)
library(ggplot2)
```

## Great resources

- [R for data science](http://r4ds.had.co.nz/): workflow, dplyr and ggplot basics
- [Stat 545](http://stat545.com/topics.html): Fantastic tutorials related to data exploration and analysis.
- [Advanced R](http://adv-r.had.co.nz/): data structure, writing better functions, R programming in general
- [Sytle guide](http://style.tidyverse.org/): write organized, readable code!


## Install packages

- Use command `install.packages("package-name")` (Don't forget the quotes!)
- We will use the package collection [`tidyverse`](https://www.tidyverse.org/packages)
- `tidyvserse` includes
  - [ggplot](http://ggplot2.tidyverse.org/)
  - [dplyr](http://dplyr.tidyverse.org/)
- Load packages using `library(package-name)` (No quotes!)

## Some quick tips

- `<-` assignment operator (Alt/Option -)
- Ctrl + Enter to send line to console (you can highlight multiple lines)
- Ctrl + Shift + Enter runs the entire script
- Many more ways to send code, see: Rstudio menu > code
- Remember to use **tab completion**!
- Get help using `?functionName` or `??partialFunctionName`
- Other keyboard short cuts: Rstudio menu > Tools > keyboard

## Importing data
- Excel files: use the package `readxl`
- Various `read_...` functions 

## Quick exploration
- What type of data do we have?
- Categorical, numerical, boolean, character

## Exploration
```{r echo = TRUE, eval=FALSE}
# Import data set
bio_df <- readxl::read_xlsx('data/SSMI Data Sheet Spray Trial 1.xlsx')

# Taking a quick look at our data is easy
bio_df
View(bio_df)
```

## Exploration
```{r echo = F, eval=T}
# Import data set
bio_df <- readxl::read_xlsx('data/SSMI Data Sheet Spray Trial 1.xlsx')

# Taking a quick look at our data is easy
bio_df
```

## dplyr basics
- The [pipe](http://r4ds.had.co.nz/pipes.html) %>% (Ctrl + Shift + M)
- Pick observations by their values `filter()`.
- Reorder the rows `arrange()`.
- Pick variables by their names `select()`.
- Create new variables with functions of existing variables `mutate()`.
- Collapse many values down to a single summary `summarise()`.

## Selecting columns
```{r echo = T}
# Selecting columns
bio_df %>% 
  select(Mating_status:Eggs) %>% 
  head()
```

## Filtering
```{r echo = T}
# Filtering 
bio_df %>% 
  filter(Sex == "F")
```
- [List of logic operators](https://www.statmethods.net/management/operators.html)


## Grouping and summarize
```{r echo = T}
bio_df %>% 
  select(Mating_status, Survival) %>% 
  group_by(Mating_status) %>% 
  summarize(mean_sur = mean(Survival))
```

## Factors
[Read more about factors](r3ds.had.co.nz/factors.html)
```{r echo = T}
sex_levels <- c("M", "F")
factor(bio_df$Sex, sex_levels)
```

## Factors
```{r echo = TRUE, warning=FALSE}
# Make sure the data is in a useable form
bio_df <- bio_df %>% 
  mutate(Sex = factor(Sex)) %>% 
  mutate(Population = factor(Population)) %>% 
  mutate(Cage_Number = factor(Cage_Number)) %>% 
  mutate(Treatment = factor(Treatment)) %>% 
  mutate(Mating_status = factor(Mating_status)) %>% 
  mutate(Eggs = as.numeric(Eggs)) 

head(bio_df)
```

## Plotting
```{r echo =TRUE}
# Scatter plot matrix
plot(bio_df[, 5:10])
```

## Box plot
```{r echo =TRUE}
# Scatter plot matrix
plot(x = bio_df$Sex, y = bio_df$Survival)
```

## ggplot
```{r, echo = TRUE, fig.height=3}
bio_df %>% ggplot(aes(x = Sex, y = Survival)) +
  geom_boxplot()
```
- Note ggplot grammar is `ggplot()` **+** `geom_somethingsomething()`. This
happened due to historical reasons.


## Why ggplot
```{r, echo = TRUE}
dayplot <- bio_df %>% ggplot(aes(x = Day, y = Survival)) 

dayplot + geom_point()
```

## Why ggplot
```{r, echo = TRUE, eval=FALSE}
# Data is already mapped, adding extra elements
dayplot + 
  geom_point(aes(color = Mating_status))
```

## Why ggplot
```{r, echo = TRUE}
dayplot + 
  geom_point(aes(color = Mating_status))
```

## Why ggplot
```{r, echo = TRUE, fig.height=3.5, message=FALSE}
dayplot + 
  geom_point(aes(color = Sex)) +
  facet_wrap(Sex ~ Mating_status)
```


## Linear regression
```{r}
linear_mod <- lm(Survival ~ Day, data = bio_df)
summary(linear_mod)
```

##
We can quickly see that it's all nonsense
```{r}
dayplot + 
  geom_point() +
  stat_smooth(method = "lm", color = "red")
```

## Exercise Time


